---
title: "Mapping Visualization Essay"
author: "Laura Crossley"
date: "4/10/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE, message = FALSE, quiet = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, message = FALSE, quiet = TRUE}
library(sf)
library(leaflet)
library(tidyverse)
library(ggmap)
```


I am interested in urban Native American history in the late nineteenth century. This is something I have only studied on the level of individual cities and individual people (and usually in the context of traveling rather than residing), so I was very curious what I would find. I quickly discovered some difficulties in creating visualizations that would allow me to see the Native presence in urban areas, though.^[Note that this exploration does not begin to deal with questions of the "authenticity" of census data.] Creating a basic map of the number of people per county who identified their race as Indian in the 1890 census gives the map below (note that this census data counts only Native people living outside of reservations). This map gives the impression that the Native presence outside the far West is negligible beyond a few pockets (Thurston County, NE; Knox County, NE; Neshoba County, MS; Swain County, NC; Emmet County, MI; Brown County, WI; Outagamie County, WI), none of which are locations of major urban centers.^[https://www.census.gov/population/www/documentation/twps0027/tab12.txt I only plotted the top 25 urban centers of the eastern half of the United States because that is my main geographic focus for the rest of this visualization essay. San Francisco and Denver, the only two far western cities in the top 25, were not plotted.] With the lightest color of the map representing every number from 0 to 500, there are a lot of Native people, including urban-dwelling Native people, rendered invisible by this map.


```{r echo = FALSE, message = FALSE, quiet = TRUE, results = "hide"}

city <- c("New York", "Chicago", "Philadelphia", "Brooklyn", "St. Louis", "Boston", "Baltimore", "Cincinnati", "Cleveland", "Buffalo", "New Orleans", "Pittsburgh", "Washington, DC", "Detroit", "Milwaukee", "Newark", "Minneapolis", "Jersey City", "Louisville", "Omaha", "Rochester", "St. Paul", "Kansas City, MO", "Providence", "Indianapolis")

geocodes <- geocode(city)

cities_25 = data.frame(city, geocodes)

counties_1890 <- st_read("US_county_1890_conflated.shp", stringsAsFactors = FALSE)

census_1890 <- read_csv("nhgis0008_csv/nhgis0008_ds27_1890_county.csv")

map_1890 <- counties_1890 %>% 
  filter(!STATENAM %in% c("Alaska Territory", "Hawaii Territory")) %>% 
  st_transform(4326)

counties_with_data_1890 <- map_1890 %>% 
  left_join(census_1890, by = "GISJOIN")

census_pal <- colorBin("Purples",
                       domain = counties_with_data_1890$AVF010,
                       bins = 7, na.color = "#808080")
```
```{r echo = FALSE, message = FALSE, quiet = TRUE}
leaflet(counties_with_data_1890) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~census_pal(AVF010),
              fillOpacity = 1,
              color = "black",
              opacity = 1,
              weight = 1,
              label = ~paste(NHGISNAM, ": ", AVF010)) %>% 
  addMarkers(~cities_25$lon, ~cities_25$lat, label = ~as.character(city)) %>%
  addLegend("bottomright", pal = census_pal,
            values = ~AVF010, opacity = 1,
            title = "1890 Indian Population")
```


Thinking that taking out the places with the largest populations should reveal a finer grained view of population levels, I removed the states/territories west of the Dakotas, Nebraska, Kansas, and Texas.^[The western half of the map is also misleading because the census data does not include Native people living on reservations, and most reservations are located in the West. I also filtered out Thurston County, NE, because the county's Native population of 1,898 was almost 1,000 more than the next highest number for any other county, making a finer grained view impossible.] This did not change the map much, though, because almost every county still fell within a range of 0-200 people who identified their race as Indian. I have learned to be skeptical of quantifying Native Americans within the population not only because of the ways that quantification has been used as a colonial tool against the sovereignty of Native nations but also because it is often not possible to speak in terms of large numbers of people when studying Native American history. Historical Native experiences should not be discounted just because of small numbers, though, especially because discounting urban Native Americans perpetuates the settler myth that Native people and cities are mutually exclusive, that one represents the past and the other the modern/the future.^[This is also why I chose not to create maps of Native people as percentages of the total population.]

For this reason, I decided that a better way to visualize Native presence would be to filter out counties where zero people identified their race as Indian. This map makes it immediately clear that there is still a Native presence in the eastern half of the United States, where settler memory has attempted to erase their continued presence. It is also now clear that there were people identifying as Indian in the counties of all of the top 25 most populous cities of the eastern half.^[https://www.census.gov/population/www/documentation/twps0027/tab12.txt These are the top 25 most populous cities in the country after taking out San Francisco and Denver (because they are excluded from the current map).] Only having county level data to work from, it is possible that some of these numbers come from people living in non-urban parts of the same county, but this is not likely the case for every person in every county, especially given the small size of many of these counties (particularly those of the largest cities). Plus, living in a rural area very close to a major city involves a certain level of exposure to urban living and the effects of urbanization.

I was most surprised to see the 258 figure for Philadelphia County, the largest for any of the major cities above. From what I can find, there has been little research on Native people in nineteenth- or twentieth-century Philadelphia, so I do not have an explanation for this.^[Interestingly, as the third map shows, the number for Philadelphia in 1880 is only 30. I would need to do some future research to see how the number jumped by 228 to 258 by 1890. I am not sure if there is an error in the data or if there is something else going on.]


```{r echo = FALSE, message = FALSE, quiet = TRUE}


counties_with_data_1890 <- map_1890 %>% 
  left_join(census_1890, by = "GISJOIN") %>%
  filter(!STATENAM %in% c("California", "Arizona Territory", "New Mexico Territory", "Utah Territory", "Colorado", "Wyoming Territory", "Wyoming", "Washington", "Oregon", "Montana", "Idaho", "Nevada")) %>%
  filter(NHGISNAM != "Thurston")

counties_with_data_1890_2 <- counties_with_data_1890 %>%
  filter(AVF010 > 0)

census_pal <- colorBin("Purples",
                       domain = counties_with_data_1890_2$AVF010,
                       bins = 5, na.color = "#808080")

leaflet(counties_with_data_1890_2) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~census_pal(AVF010),
              fillOpacity = .8,
              color = "black",
              opacity = 1,
              weight = 1,
              label = ~paste(NHGISNAM, ": ", AVF010)) %>% 
  addMarkers(~cities_25$lon, ~cities_25$lat, label = ~as.character(city)) %>%
  addLegend("bottomright", pal = census_pal,
            values = ~AVF010, opacity = 1,
            title = "1890 Indian Population")

```


Finally, I created a map of the change from 1880 to 1890 in the number of people who identified as Indian.^[I opted not to map percent changes because the numbers of people involved are generally very small. This time I filtered to only map the counties with at least 1 person identifying as Indian in at least one of the two years.] Looking at the counties of 1890's top 25 most populous cities in the eastern half of the United States reveals fifteen where the number increased versus nine where the number decreased (Washington, DC, is not included in the 1880 census data, so a comparison is not possible). Two of these counties, Allegheny (where Pittsburgh is located) and Milwaukee, show an increase from zero people who identified as Indian in 1880. All of this is evidence to support the argument that Native American urbanization was taking place in the late-nineteenth century, even if not at the same scale or percentages as the country as a whole.


```{r echo = FALSE, message = FALSE, quiet = TRUE, results = "hide"}
census_1880 <- read_csv("nhgis0005_csv/nhgis0005_ds23_1880_county.csv")

combined_1880_1890 <- counties_with_data_1890 %>% 
  left_join(census_1880, by = "GISJOIN") %>%
  filter(AVF010 > 0 | APP004 > 0) %>%
  mutate(change = AVF010-APP004)

census_pal <- colorBin("PuOr",
                       domain = combined_1880_1890$change,
                       bins = 5, na.color = "#808080")
```
```{r echo = FALSE, message = FALSE, quiet = TRUE}
leaflet(combined_1880_1890) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~census_pal(change),
              fillOpacity = .8,
              color = "black",
              opacity = 1,
              weight = 1,
              label = ~paste(NHGISNAM, ": ", APP004, " to ", AVF010, "(change of ", change, ")")) %>%
  addMarkers(~cities_25$lon, ~cities_25$lat, label = ~as.character(city)) %>%
  addLegend("bottomright", pal = census_pal,
            values = ~change, opacity = 1,
            title = "Change in Indian Population from 1880 to 1890")
```
